---
title: "The topology of software risk in scientific models"
subtitle: "5. Dynamic profiling validation"
author: "Arnald Puy"
header-includes:
  - \usepackage[font=footnotesize]{caption}
  - \usepackage{dirtytalk}
  - \usepackage{booktabs}
  - \usepackage{tabulary}
  - \usepackage{enumitem}
  - \usepackage{lmodern}
  - \usepackage{amsmath}
  - \usepackage{mathtools}
  - \usepackage[T1]{fontenc}
  - \usepackage{tikz}
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
    keep_tex: true
  word_document:
    toc: no
    toc_depth: '2'
  html_document:
    keep_md: true
link-citations: yes
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "pdf", cache = TRUE)
```

\newpage

# Preliminary

This document layers a dynamic-profiling validation on top of the static
path-level software-risk framework built in
`code_hydrological_models.Rmd`. For each model with available runtime
traces (currently ORCHIDEE and PCR-GLOBWB; VIC and HYPE expected), it
joins the static call graph to the per-edge runtime call counts and runs
the five analyses pre-registered in `dynamic_validation/pre_registration.md`:

1. **Path coverage.** Are statically high-risk paths preferentially
   exercised at runtime?
2. **Rank correlation.** Among exercised paths, does $P_k$ order
   runtime intensity $E_k^{sum}$ correctly?
3. **Reverse predictive check.** Restricting to runtime-heavy paths,
   does static $P_k$ still discriminate on the underlying node-level
   metrics?
4. **Matched-frequency contrast.** Does static topology add information
   beyond execution frequency, when high- and low-$P_k$ paths are
   matched on $E_k^{sum}$, length and statement count?
5. **Dormant high-risk paths.** Which statically high-risk paths are
   not (fully) exercised in the chosen configuration?

The individual analyses also live as standalone scripts under
`dynamic_validation/analysis/` for ad-hoc re-runs and as a development
view. This `.Rmd` is the canonical reference: every number reported
in `dynamic_validation/summary.md` and
`dynamic_validation/draft_results_paragraph.md` is reproduced here.

```{r packages, warning=FALSE, message=FALSE, results = "hide", cache = FALSE}

# PRELIMINARY FUNCTIONS ########################################################
################################################################################

sensobol::load_packages(c("data.table", "tidyverse", "openxlsx", "scales",
                          "cowplot", "readxl", "ggrepel", "tidytext", "here",
                          "tidygraph", "igraph", "foreach", "parallel", "ggraph",
                          "tools", "purrr", "sensobol", "benchmarkme", "softwareRisk",
                          "boot", "MatchIt", "effsize"))

# Set seed ---------------------------------------------------------------------

seed <- 123
set.seed(seed)
```

\newpage

# Helpers and model registry

This section bundles the data loaders, the model registry (which models
have runtime data and which language each was profiled in), and the
per-path helpers used throughout. Mirrors `dynamic_validation/analysis/00_setup.R`.

```{r helpers}

# MODEL REGISTRY ###############################################################
# Append rows here when VIC and HYPE arrive. `model` matches keys in
# full_paths_df.xlsx / full_node_df.xlsx; `raw_subdir` is the folder under
# dynamic_validation/.

model_registry <- data.table(
  model      = c("ORCHIDEE", "PCR-GLOBWB"),
  raw_subdir = c("ORCHIDEE", "PCR"),
  lang       = c("fortran",  "python")
)

repo_root <- getwd()
dv_dir    <- file.path(repo_root, "dynamic_validation")
raw_dir   <- function(model) file.path(dv_dir, model)
out_dir   <- file.path(dv_dir, "results", "per_model")
syn_dir   <- file.path(dv_dir, "results", "synthesis")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(syn_dir, recursive = TRUE, showWarnings = FALSE)

available_models <- function() {
  reg <- copy(model_registry)
  reg[, available := vapply(raw_subdir, function(s) {
    file.exists(file.path(raw_dir(s), "runtime_callgraph_edges.csv"))
  }, logical(1))]
  reg[available == TRUE]
}

# NAME NORMALISATION ###########################################################
# Both static and runtime caller / callee names are lowercased and stripped of
# a leading class qualifier. Adjust here if model-specific quirks emerge
# (Fortran name mangling, Python decorators, etc.).

norm_fun_name <- function(x) {
  out <- tolower(trimws(as.character(x)))
  out <- sub(".*\\.", "", out)
  out
}

# STATIC SIDE LOADERS ##########################################################
# Read the canonical outputs of code_hydrological_models.R.

load_full_paths_df <- function() {
  f <- file.path(repo_root, "full_paths_df.xlsx")
  if (!file.exists(f)) stop("Missing ", f,
    " - run code_hydrological_models.R first.")
  as.data.table(read.xlsx(f))
}

load_full_nodes_df <- function() {
  f <- file.path(repo_root, "full_nodes_df.xlsx")
  if (!file.exists(f)) stop("Missing ", f,
    " - run code_hydrological_models.R first.")
  as.data.table(read.xlsx(f))
}

load_full_node_df <- function() {
  f <- file.path(repo_root, "full_node_df.xlsx")
  if (!file.exists(f)) stop("Missing ", f,
    " - run code_hydrological_models.R first.")
  as.data.table(read.xlsx(f))
}

# Per-path uncertainty summary - prefers the slim full_ua_summary.csv
# (produced by dynamic_validation/analysis/00b_extract_ua_summary.R; ~8.7 MB)
# and falls back to the raw 1.78 GB full_ua_df.csv with an sd approximated
# from quantiles.

load_uncertainty_summary <- function(models = NULL) {
  slim <- file.path(repo_root, "full_ua_summary.csv")
  raw  <- file.path(repo_root, "full_ua_df.csv")
  if (file.exists(slim)) {
    ua <- fread(slim)
  } else if (file.exists(raw)) {
    cols <- c("model", "path_id", "path_nodes", "path_str", "hops",
              "P_k_min", "P_k_mean", "P_k_max",
              "P_k_q025", "P_k_q50", "P_k_q975")
    ua <- fread(raw, select = cols)
    ua[, P_k_sd := (P_k_q975 - P_k_q025) / (2 * 1.96)]
    ua[, CV_k   := P_k_sd / pmax(P_k_mean, 1e-12)]
  } else {
    stop("Missing both ", slim, " and ", raw,
         " - run code_hydrological_models.R UA section, then ",
         "dynamic_validation/analysis/00b_extract_ua_summary.R.")
  }
  if (!is.null(models)) ua <- ua[model %in% models]
  if (!"risk_form" %in% names(ua)) ua[, risk_form := "additive"]
  ua[]
}

load_static_edges <- function(model_label, lang) {
  fn <- sprintf("call_metrics_%s_%s.csv", lang, model_label)
  f  <- file.path(repo_root, "datasets", "call_metrics", fn)
  if (!file.exists(f)) stop("Missing ", f)
  dt <- fread(f)
  setnames(dt, c("function", "call"), c("caller", "callee"), skip_absent = TRUE)
  dt[, `:=`(
    caller_norm = norm_fun_name(caller),
    callee_norm = norm_fun_name(callee),
    model       = model_label,
    lang        = lang
  )]
  unique(dt[, .(caller_norm, callee_norm, model, lang)])
}

# DYNAMIC SIDE LOADER ##########################################################
# runtime_callgraph_edges.csv schema (as of 2026-05-17): model, lang, caller,
# callee, caller_file, caller_line, callee_file_hint, callee_kind,
# runtime_calls. Name normalisation can collapse multiple raw edges into the
# same (caller_norm, callee_norm); sum the runtime_calls to preserve intensity.

load_runtime_edges <- function(model_label, raw_subdir) {
  f <- file.path(raw_dir(raw_subdir), "runtime_callgraph_edges.csv")
  if (!file.exists(f)) stop("Missing ", f)
  dt <- fread(f)
  dt[, `:=`(
    caller_norm = norm_fun_name(caller),
    callee_norm = norm_fun_name(callee)
  )]
  if (!"runtime_calls" %in% names(dt)) dt[, runtime_calls := NA_integer_]
  dt[, .(runtime_calls = sum(runtime_calls, na.rm = TRUE)),
     by = .(caller_norm, callee_norm)]
}

# PER-PATH HELPERS #############################################################
# path_nodes in full_paths_df is a delimited string after the xlsx round-trip.
# split_path_nodes returns the ordered node sequence; path_to_edges turns it
# into a per-edge table; aggregate_Ek collapses runtime counts along a path.

split_path_nodes <- function(path_nodes_str) {
  for (sep in c(" -> ", ",", "|")) {
    if (any(grepl(sep, path_nodes_str, fixed = TRUE))) {
      return(strsplit(path_nodes_str, sep, fixed = TRUE))
    }
  }
  strsplit(path_nodes_str, "\\s+")
}

path_to_edges <- function(node_seq) {
  if (length(node_seq) < 2L) {
    return(data.table(caller_norm = character(), callee_norm = character()))
  }
  data.table(
    caller_norm = norm_fun_name(node_seq[-length(node_seq)]),
    callee_norm = norm_fun_name(node_seq[-1L])
  )
}

aggregate_Ek <- function(edge_dt) {
  rc <- edge_dt$runtime_calls
  rc[is.na(rc)] <- 0
  if (length(rc) == 0L) return(list(Ek_min = NA_real_, Ek_sum = NA_real_))
  list(Ek_min = min(rc), Ek_sum = sum(rc))
}

# Expand each static path into one row per (path_id, position, node) for a
# given (model, risk_form), with node-level attributes (cyclomatic_complexity,
# indeg, btw, risk_score, file) attached. Used by Analyses 3, 4, 5.

expand_paths_to_nodes <- function(model_label,
                                  risk_form_label = "additive",
                                  paths_df = NULL,
                                  node_df  = NULL) {
  if (is.null(paths_df)) paths_df <- load_full_paths_df()
  if (is.null(node_df))  node_df  <- load_full_node_df()

  fp <- paths_df[model == model_label & risk_form == risk_form_label]
  fn <- node_df[model == model_label,
                .(model, name, file, cyclomatic_complexity, indeg, btw,
                  risk_score)]
  if (!nrow(fp)) return(data.table())

  per_path <- rbindlist(lapply(seq_len(nrow(fp)), function(i) {
    nodes <- split_path_nodes(fp$path_nodes[i])[[1]]
    if (!length(nodes)) return(NULL)
    data.table(model = model_label, path_id = fp$path_id[i],
               pos = seq_along(nodes), name = nodes)
  }), use.names = TRUE, fill = TRUE)

  merge(per_path, fn, by = c("model", "name"), all.x = TRUE)
}

cat("[setup] models with runtime data:\n")
print(available_models())
```

\newpage

# Pre-registration

The decisions in `dynamic_validation/pre_registration.md` were locked
before any analysis was run against the dynamic data. Highlights
relevant to interpreting the numbers below:

- **Path-level aggregation of $E_k$.** Primary aggregator: `sum`
  (originally `min`; switched on 2026-05-17 because only 38 of 3,152
  ORCHIDEE paths had every edge exercised at least once, making `min`
  collapse to a near-binary indicator informationally identical to
  `all_exercised`). `min` is reported alongside as a robustness check
  in Analysis 2.
- **High-risk stratum:** top decile of $P_k$ AND $\text{CV}(P_k)$ below
  the model-specific median CV. Matches the bug-injection filter in
  the manuscript.
- **Low-risk stratum:** bottom decile of $P_k$ (no CV filter).
- **Effect-size convention:** common-language $A$ (Vargha & Delaney),
  matching Table S2 of the manuscript.
- **Predictive threshold:** $A > 0.55$ at $p < 0.05$ within a given
  comparison; the global decision rule for "framework predicts at
  runtime" additionally requires Spearman $\rho > 0$ with bootstrap CI
  strictly above zero, in at least three of the four target models.

```{r preregistration}

# Decision parameters used throughout - all locked in pre_registration.md.

TOP_DECILE        <- 0.90
BOTTOM_DECILE     <- 0.10
RISK_FORM_PRIMARY <- "additive"
EK_PRIMARY        <- "Ek_sum"   # see deviation log 2026-05-17
EK_ROBUSTNESS     <- "Ek_min"
N_BOOT            <- 1000L
A_THRESHOLD       <- 0.55
P_THRESHOLD       <- 0.05
TOP_K_FUNCTIONS   <- 3L
MIN_PATHS_PER_BIN <- 20L  # within-length deciles require at least this many

color_strata <- c("top_decile" = "#C65D09", "bottom_decile" = "#3B5B92")
```

\newpage

# Static-dynamic join

For each model with available runtime data, build a per-path joined
table containing:

- the static path identifiers and risk summary (`P_k_q50`, `CV_k`,
  `hops`),
- per-path counts of edges exercised at runtime under the looser
  ("any") and stricter ("all") definitions,
- per-path runtime intensity under both aggregators (`Ek_min`,
  `Ek_sum`), with unmatched edges contributing zero calls.

```{r join_static_dynamic, results="hold"}

# JOIN STATIC + DYNAMIC ########################################################
################################################################################

ua_summary  <- load_uncertainty_summary(models = available_models()$model)
paths_df    <- load_full_paths_df()
node_df     <- load_full_node_df()
nodes_df    <- load_full_nodes_df()

join_one_model <- function(model_label, raw_subdir, lang) {

  static_edges  <- load_static_edges(model_label, lang)
  runtime_edges <- load_runtime_edges(model_label, raw_subdir)

  matched_edges <- merge(static_edges, runtime_edges,
                         by = c("caller_norm", "callee_norm"),
                         all.x = TRUE)
  matched_edges[, exercised := !is.na(runtime_calls)]
  setkey(matched_edges, caller_norm, callee_norm)

  paths <- ua_summary[model == model_label]

  per_path <- paths[, {
    node_seq <- split_path_nodes(path_nodes)[[1]]
    edges    <- path_to_edges(node_seq)
    if (nrow(edges) == 0L) {
      list(n_edges = 0L, n_edges_static = 0L, n_edges_exercised = 0L,
           frac_exercised = NA_real_, any_exercised = NA, all_exercised = NA,
           Ek_min = NA_real_, Ek_sum = NA_real_)
    } else {
      joined <- matched_edges[edges, on = c("caller_norm", "callee_norm")]
      ex     <- !is.na(joined$exercised) & joined$exercised
      Ek     <- aggregate_Ek(joined)
      list(n_edges = nrow(edges),
           n_edges_static = sum(!is.na(joined$model)),
           n_edges_exercised = sum(ex),
           frac_exercised = sum(ex) / nrow(edges),
           any_exercised = any(ex),
           all_exercised = all(ex),
           Ek_min = Ek$Ek_min,
           Ek_sum = Ek$Ek_sum)
    }
  }, by = .(model, path_id, risk_form, hops,
            P_k_min, P_k_mean, P_k_max, P_k_q025, P_k_q50, P_k_q975, CV_k)]

  out_f <- file.path(out_dir, paste0(model_label, "_paths_joined.csv"))
  fwrite(per_path, out_f)

  diag <- data.table(
    model            = model_label,
    lang             = lang,
    static_edges     = nrow(static_edges),
    runtime_edges    = nrow(runtime_edges),
    matched_edges    = sum(matched_edges$exercised),
    runtime_coverage_pct =
      100 * sum(matched_edges$exercised) / nrow(static_edges)
  )
  fwrite(diag, file.path(out_dir, paste0(model_label, "_edge_diagnostics.csv")))
  list(paths = per_path, diag = diag)
}

reg <- available_models()
join_results <- lapply(seq_len(nrow(reg)), function(i) {
  join_one_model(reg$model[i], reg$raw_subdir[i], reg$lang[i])
})
names(join_results) <- reg$model

# Pool joined tables for downstream analysis
all_joined <- rbindlist(lapply(join_results, `[[`, "paths"))

# Edge-level coverage diagnostics
edge_diag <- rbindlist(lapply(join_results, `[[`, "diag"))
print(edge_diag)
```

The edge-level diagnostics quantify how much of each model's static
call graph the runtime trace actually reached. They are a precondition
for every downstream coverage claim: if the matched-edge fraction is
small, the dormancy numbers below are bounded by it.

\newpage

# Analysis 1: Path coverage

## Unstratified contrast

Top decile vs bottom decile of $P_k$, three coverage metrics:

- `frac_any`: at least one edge of the path was exercised (loose).
- `frac_edge`: mean fraction of edges per path that were exercised
  (length-insensitive).
- `frac_all`: every edge of the path was exercised (strict).

```{r analysis1_unstratified}

# ANALYSIS 1 - PATH COVERAGE (UNSTRATIFIED) ####################################
################################################################################

coverage_dt <- all_joined[
  risk_form == RISK_FORM_PRIMARY & !is.na(P_k_q50),
  {
    qhi    <- quantile(P_k_q50, TOP_DECILE,    na.rm = TRUE)
    qlo    <- quantile(P_k_q50, BOTTOM_DECILE, na.rm = TRUE)
    cv_med <- median(CV_k, na.rm = TRUE)
    hi_idx <- P_k_q50 >= qhi & CV_k < cv_med
    lo_idx <- P_k_q50 <= qlo
    rbind(
      data.table(stratum = "top_decile",
                 n_paths = sum(hi_idx),
                 frac_any = mean(any_exercised[hi_idx], na.rm = TRUE),
                 frac_all = mean(all_exercised[hi_idx], na.rm = TRUE),
                 frac_edge = mean(frac_exercised[hi_idx], na.rm = TRUE),
                 median_hops = as.numeric(median(hops[hi_idx], na.rm = TRUE))),
      data.table(stratum = "bottom_decile",
                 n_paths = sum(lo_idx),
                 frac_any = mean(any_exercised[lo_idx], na.rm = TRUE),
                 frac_all = mean(all_exercised[lo_idx], na.rm = TRUE),
                 frac_edge = mean(frac_exercised[lo_idx], na.rm = TRUE),
                 median_hops = as.numeric(median(hops[lo_idx], na.rm = TRUE)))
    )
  },
  by = .(model, risk_form)
]

fwrite(coverage_dt, file.path(syn_dir, "01_coverage_table.csv"))
print(coverage_dt)
```

```{r analysis1_unstratified_plot, fig.width=8, fig.height=4.5, fig.cap="Analysis 1 unstratified: fraction of paths exercised at runtime by static-risk stratum, under three coverage metrics. Note the median-hops column in the table - top-decile paths are substantially longer than bottom-decile paths, which inflates the `any` contrast."}

coverage_long <- melt(
  coverage_dt,
  id.vars       = c("model", "risk_form", "stratum", "n_paths", "median_hops"),
  measure.vars  = c("frac_any", "frac_edge", "frac_all"),
  variable.name = "metric",
  value.name    = "value"
)
coverage_long[, metric := factor(metric,
  levels = c("frac_any", "frac_edge", "frac_all"),
  labels = c("any edge", "mean fraction of edges", "all edges"))]

ggplot(coverage_long,
       aes(x = model, y = value, fill = stratum)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  facet_wrap(~ metric) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  scale_fill_manual(values = color_strata,
                    labels = c("bottom decile P_k", "top decile P_k")) +
  labs(x = NULL, y = "coverage", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

The unstratified contrast is large but partly driven by the strong
correlation between $P_k$ and path length under the saturating-OR
aggregator (Eq. 5 of the manuscript). The length-controlled version
follows.

## Within-length deciles (Analysis 1b)

For each path-length bin separately, take the top vs bottom decile of
$P_k$ within the bin and compare runtime coverage. Length is held
constant by construction; the within-bin Wilcoxon test plus the
common-language effect size $A$ summarise discrimination.

```{r analysis1b_within_length}

# ANALYSIS 1b - WITHIN-LENGTH DECILES + WILCOXON ###############################
################################################################################

within_length_dt <- all_joined[risk_form == RISK_FORM_PRIMARY & !is.na(P_k_q50)]
within_length_dt[, hops_bin := cut(hops,
                                   breaks = c(0, 1, 2, 4, 7, Inf),
                                   labels = c("1", "2", "3-4", "5-7", "8+"),
                                   right  = TRUE)]

within_decile_dt <- within_length_dt[
  ,
  {
    if (.N < MIN_PATHS_PER_BIN || length(unique(P_k_q50)) < 2L) {
      data.table()
    } else {
      qhi <- quantile(P_k_q50, TOP_DECILE,    na.rm = TRUE)
      qlo <- quantile(P_k_q50, BOTTOM_DECILE, na.rm = TRUE)
      top <- P_k_q50 >= qhi
      bot <- P_k_q50 <= qlo
      rbind(
        data.table(stratum = "top_decile",
                   n_paths = sum(top),
                   frac_any  = mean(any_exercised[top],  na.rm = TRUE),
                   frac_edge = mean(frac_exercised[top], na.rm = TRUE),
                   frac_all  = mean(all_exercised[top],  na.rm = TRUE)),
        data.table(stratum = "bottom_decile",
                   n_paths = sum(bot),
                   frac_any  = mean(any_exercised[bot],  na.rm = TRUE),
                   frac_edge = mean(frac_exercised[bot], na.rm = TRUE),
                   frac_all  = mean(all_exercised[bot],  na.rm = TRUE))
      )
    }
  },
  by = .(model, risk_form, hops_bin)
]

within_test_dt <- within_length_dt[
  ,
  {
    if (.N < MIN_PATHS_PER_BIN || length(unique(P_k_q50)) < 2L) {
      data.table()
    } else {
      qhi <- quantile(P_k_q50, TOP_DECILE,    na.rm = TRUE)
      qlo <- quantile(P_k_q50, BOTTOM_DECILE, na.rm = TRUE)
      x_top <- frac_exercised[P_k_q50 >= qhi]
      x_bot <- frac_exercised[P_k_q50 <= qlo]
      if (length(x_top) < 3L || length(x_bot) < 3L) {
        data.table()
      } else {
        w <- suppressWarnings(wilcox.test(x_top, x_bot))
        A <- tryCatch(effsize::VD.A(x_top, x_bot)$estimate,
                      error = function(e) NA_real_)
        data.table(n_top = length(x_top), n_bot = length(x_bot),
                   median_top = median(x_top, na.rm = TRUE),
                   median_bot = median(x_bot, na.rm = TRUE),
                   wilcox_p = w$p.value, effect_A = A)
      }
    }
  },
  by = .(model, risk_form, hops_bin)
]
within_test_dt[, signif := fcase(
  is.na(wilcox_p), "",
  wilcox_p < 0.001, "***",
  wilcox_p < 0.01,  "**",
  wilcox_p < 0.05,  "*",
  default = "ns"
)]

fwrite(within_decile_dt, file.path(syn_dir, "01b_within_length_deciles.csv"))
fwrite(within_test_dt,   file.path(syn_dir, "01b_within_length_tests.csv"))

print(within_decile_dt)
print(within_test_dt)
```

```{r analysis1b_plot, fig.width=8, fig.height=5, fig.cap="Analysis 1b: mean fraction of edges exercised per path, by hops bin and static-risk stratum. The decile cuts are taken WITHIN each hops bin so the top-vs-bottom comparison is length-controlled. For paths of 3 or more hops the framework discriminates strongly; for hops 1 the relationship reverses, consistent with the framework's design as a path-level integrator over multi-node chains."}

if (nrow(within_decile_dt)) {
  ggplot(within_decile_dt,
         aes(x = hops_bin, y = frac_edge, fill = stratum)) +
    geom_col(position = position_dodge(0.7), width = 0.6) +
    facet_grid(risk_form ~ model) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                       limits = c(0, 1)) +
    scale_fill_manual(values = color_strata,
                      labels = c("bottom decile", "top decile")) +
    labs(x = "path length (hops)", y = "mean fraction of edges exercised",
         fill = NULL) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
}
```

\newpage

# Analysis 2: Rank correlation between $P_k$ and $E_k$

Among paths that runtime actually exercised, does the static risk
score $P_k$ rank-order their runtime intensity? Spearman $\rho$ on
$E_k^{sum}$ (primary) and on $E_k^{min}$ (robustness; see deviation
log). Bootstrap percentile 95 % CI with $N=$ `r N_BOOT` resamples.

```{r analysis2_rank_correlation}

# ANALYSIS 2 - RANK CORRELATION P_k vs E_k #####################################
################################################################################

corr_results <- rbindlist(lapply(reg$model, function(m) {
  dt <- all_joined[model == m & risk_form == RISK_FORM_PRIMARY &
                   any_exercised == TRUE &
                   !is.na(P_k_q50) & !is.na(get(EK_PRIMARY))]
  if (!nrow(dt) || all(is.na(dt[[EK_PRIMARY]]))) return(NULL)

  rho <- suppressWarnings(cor(dt$P_k_q50, dt[[EK_PRIMARY]], method = "spearman"))
  tau <- suppressWarnings(cor(dt$P_k_q50, dt[[EK_PRIMARY]], method = "kendall"))
  rho_robust <- suppressWarnings(
    cor(dt$P_k_q50, dt[[EK_ROBUSTNESS]], method = "spearman"))

  boot_rho <- boot::boot(
    data      = dt,
    statistic = function(d, i) suppressWarnings(
      cor(d$P_k_q50[i], d[[EK_PRIMARY]][i], method = "spearman")),
    R = N_BOOT
  )
  ci <- tryCatch(
    boot::boot.ci(boot_rho, type = "perc")$percent[1, c(4, 5)],
    error = function(e) c(NA_real_, NA_real_)
  )

  data.table(model = m, n_paths = nrow(dt),
             spearman_sum = rho, kendall_sum = tau,
             rho_sum_ci_lo = ci[1], rho_sum_ci_hi = ci[2],
             spearman_min = rho_robust)
}), fill = TRUE)

fwrite(corr_results, file.path(syn_dir, "02_rank_correlation.csv"))
print(corr_results)
```

```{r analysis2_scatters, fig.width=8, fig.height=4, fig.cap="Analysis 2: P_k against log10(E_k^sum + 1) for exercised paths in each model. Spearman rho is annotated per panel; loess in orange."}

scatter_panels <- lapply(reg$model, function(m) {
  dt <- all_joined[model == m & risk_form == RISK_FORM_PRIMARY &
                   any_exercised == TRUE & !is.na(get(EK_PRIMARY))]
  if (!nrow(dt)) return(NULL)
  rho_row <- corr_results[model == m]
  ggplot(dt, aes(x = P_k_q50, y = log10(get(EK_PRIMARY) + 1))) +
    geom_point(alpha = 0.4, size = 0.8) +
    geom_smooth(method = "loess", se = FALSE, colour = "#C65D09") +
    labs(title = sprintf("%s (rho_sum = %.2f [%.2f, %.2f], n = %d)",
                         m, rho_row$spearman_sum,
                         rho_row$rho_sum_ci_lo, rho_row$rho_sum_ci_hi,
                         rho_row$n_paths),
         x = expression(P[k]~"(median across ensemble)"),
         y = expression(log[10]*"("*E[k]^{sum}+1*")")) +
    theme_minimal(base_size = 10)
})
scatter_panels <- scatter_panels[!vapply(scatter_panels, is.null, logical(1))]
if (length(scatter_panels)) {
  cowplot::plot_grid(plotlist = scatter_panels, nrow = 1)
}
```

\newpage

# Analysis 3: Reverse predictive check

Restrict to the top decile of paths by $E_k^{sum}$ (the runtime-heavy
paths), then within that subset compare path-mean cyclomatic
complexity, in-degree and betweenness between the top-$P_k$ and
bottom-$P_k$ halves. Two-sample Wilcoxon plus common-language $A$.
*Holding execution frequency fixed, does static topology still
discriminate?*

```{r analysis3_reverse_predictive}

# ANALYSIS 3 - REVERSE PREDICTIVE CHECK ########################################
################################################################################

TOP_RUNTIME_Q <- 0.90
TOP_P_Q       <- 0.50  # split high vs low P_k at the median, within the
                       # runtime-heavy subset
METRICS       <- c("cyclomatic_complexity", "indeg", "btw")

reverse_results <- list()
reverse_plots   <- list()

for (m in reg$model) {

  dt <- all_joined[model == m & risk_form == RISK_FORM_PRIMARY &
                   any_exercised == TRUE & !is.na(get(EK_PRIMARY))]
  if (!nrow(dt) || all(is.na(dt[[EK_PRIMARY]]))) next

  q_runtime <- quantile(dt[[EK_PRIMARY]], TOP_RUNTIME_Q, na.rm = TRUE)
  hot_paths <- dt[get(EK_PRIMARY) >= q_runtime, .(model, path_id, P_k_q50)]
  if (nrow(hot_paths) < 8L) next

  per_path_nodes <- expand_paths_to_nodes(m, RISK_FORM_PRIMARY,
                                          paths_df = paths_df,
                                          node_df  = node_df)
  per_path_metrics <- per_path_nodes[
    path_id %in% hot_paths$path_id,
    lapply(.SD, mean, na.rm = TRUE),
    by = .(model, path_id),
    .SDcols = METRICS
  ]
  merged <- merge(hot_paths, per_path_metrics, by = c("model", "path_id"))
  if (nrow(merged) < 8L) next

  q_pk <- quantile(merged$P_k_q50, TOP_P_Q)
  merged[, stratum := ifelse(P_k_q50 >= q_pk, "high_Pk", "low_Pk")]

  per_metric <- lapply(METRICS, function(mt) {
    a <- merged[stratum == "high_Pk", get(mt)]
    b <- merged[stratum == "low_Pk",  get(mt)]
    if (!length(a) || !length(b)) return(NULL)
    w <- suppressWarnings(wilcox.test(a, b))
    A <- tryCatch(effsize::VD.A(a, b)$estimate, error = function(e) NA_real_)
    data.table(model = m, metric = mt,
               n_high = length(a), n_low = length(b),
               median_high = median(a, na.rm = TRUE),
               median_low  = median(b, na.rm = TRUE),
               wilcox_p = w$p.value, effect_A = A)
  })
  reverse_results[[m]] <- rbindlist(per_metric, fill = TRUE)

  plot_dt <- melt(merged, id.vars = c("model", "path_id", "stratum"),
                  measure.vars = METRICS,
                  variable.name = "metric", value.name = "path_mean")
  reverse_plots[[m]] <- ggplot(plot_dt,
    aes(x = stratum, y = path_mean, fill = stratum)) +
    geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
    facet_wrap(~ metric, scales = "free_y", nrow = 1) +
    scale_fill_manual(values = c("high_Pk" = "#C65D09", "low_Pk" = "#3B5B92")) +
    labs(title = m, x = NULL, y = "path-mean") +
    theme_minimal(base_size = 10) + theme(legend.position = "none")
}

if (length(reverse_results)) {
  reverse_tbl <- rbindlist(reverse_results, fill = TRUE)
  fwrite(reverse_tbl, file.path(syn_dir, "03_reverse_predictive.csv"))
  print(reverse_tbl)
}
```

```{r analysis3_plot, fig.width=8, fig.height=3.5, fig.cap="Analysis 3: distribution of path-mean static metrics (cyclomatic complexity, in-degree, betweenness) between high-Pk and low-Pk strata within the runtime-heavy decile, per model."}

if (length(reverse_plots)) {
  cowplot::plot_grid(plotlist = reverse_plots, ncol = 1, align = "v")
}
```

\newpage

# Analysis 4: Matched-frequency contrast

The cleanest test of "does static topology add information beyond
execution frequency": match each top-decile high-$P_k$ path to a
bottom-decile low-$P_k$ path with comparable $E_k^{sum}$, hops and
total statement count (Mahalanobis nearest-neighbour with replacement,
`MatchIt`). Then compare static metrics on the matched set. Sample
sizes below 15 pairs are flagged as exploratory.

```{r analysis4_matched_contrast}

# ANALYSIS 4 - MATCHED-FREQUENCY CONTRAST ######################################
################################################################################

EXPLORATORY_THRESHOLD <- 15L
matched_results <- list()
diag_smd        <- list()

for (m in reg$model) {

  dt <- all_joined[model == m & risk_form == RISK_FORM_PRIMARY &
                   any_exercised == TRUE & !is.na(get(EK_PRIMARY))]
  if (!nrow(dt) || all(is.na(dt[[EK_PRIMARY]]))) next

  qhi    <- quantile(dt$P_k_q50, TOP_DECILE,    na.rm = TRUE)
  qlo    <- quantile(dt$P_k_q50, BOTTOM_DECILE, na.rm = TRUE)
  cv_med <- median(dt$CV_k, na.rm = TRUE)

  dt[, group := fifelse(P_k_q50 >= qhi & CV_k < cv_med, 1L,
                fifelse(P_k_q50 <= qlo,                 0L, NA_integer_))]
  dt_match <- dt[!is.na(group)]

  if (sum(dt_match$group == 1L) < 5L || sum(dt_match$group == 0L) < 5L) next

  per_path_nodes <- expand_paths_to_nodes(m, RISK_FORM_PRIMARY,
                                          paths_df = paths_df,
                                          node_df  = node_df)
  metric_dt <- per_path_nodes[
    path_id %in% dt_match$path_id,
    lapply(.SD, mean, na.rm = TRUE),
    by = .(path_id),
    .SDcols = METRICS
  ]
  dt_match <- merge(dt_match, metric_dt, by = "path_id", all.x = TRUE)
  dt_match[, statement_proxy := cyclomatic_complexity * hops]

  mout <- tryCatch(
    MatchIt::matchit(
      group ~ get(EK_PRIMARY) + hops + statement_proxy,
      data    = dt_match,
      method  = "nearest",
      distance = "mahalanobis",
      replace  = TRUE
    ),
    error = function(e) { warning(e$message); NULL }
  )
  if (is.null(mout)) next

  matched <- MatchIt::match.data(mout)
  n_pairs <- sum(matched$group == 1L)

  smd_pre  <- summary(mout)$sum.all[, "Std. Mean Diff."]
  smd_post <- summary(mout)$sum.matched[, "Std. Mean Diff."]
  diag_smd[[m]] <- data.table(
    model     = m,
    covariate = names(smd_post),
    smd_pre   = smd_pre[names(smd_post)],
    smd_post  = smd_post
  )

  per_metric <- lapply(METRICS, function(mt) {
    a <- matched[[mt]][matched$group == 1L]
    b <- matched[[mt]][matched$group == 0L]
    if (!length(a) || !length(b)) return(NULL)
    w <- suppressWarnings(wilcox.test(a, b))
    A <- tryCatch(effsize::VD.A(a, b)$estimate, error = function(e) NA_real_)
    data.table(model = m, metric = mt, n_pairs = n_pairs,
               median_high = median(a, na.rm = TRUE),
               median_low  = median(b, na.rm = TRUE),
               wilcox_p = w$p.value, effect_A = A,
               exploratory = n_pairs < EXPLORATORY_THRESHOLD)
  })
  matched_results[[m]] <- rbindlist(per_metric, fill = TRUE)
}

if (length(matched_results)) {
  matched_tbl <- rbindlist(matched_results, fill = TRUE)
  fwrite(matched_tbl, file.path(syn_dir, "04_matched_contrast.csv"))
  fwrite(rbindlist(diag_smd, fill = TRUE),
         file.path(syn_dir, "04_matched_smd_diagnostics.csv"))
  print(matched_tbl)
}
```

\newpage

# Analysis 5: Dormant high-risk paths

Of the within-length top-decile high-risk paths in each length bin,
how many were not (fully) exercised in the chosen runtime
configuration? *Loose* dormancy means no edge of the path fired;
*strict* dormancy means at least one edge of the path did not fire.
The within-length top-decile cut (rather than the overall top decile)
ensures the dormant set is not just long paths that mechanically have
more chances to fail.

```{r analysis5_dormant_paths}

# ANALYSIS 5 - DORMANT HIGH-RISK PATHS #########################################
################################################################################

dormant_summary   <- list()
dormant_paths_out <- list()

for (m in reg$model) {

  dt <- all_joined[model == m & risk_form == RISK_FORM_PRIMARY & !is.na(P_k_q50)]
  if (!nrow(dt)) next
  dt[, hops_bin := cut(hops,
                       breaks = c(0, 1, 2, 4, 7, Inf),
                       labels = c("1", "2", "3-4", "5-7", "8+"),
                       right  = TRUE)]

  hi <- dt[
    ,
    {
      if (.N < MIN_PATHS_PER_BIN || length(unique(P_k_q50)) < 2L) {
        data.table()
      } else {
        qhi    <- quantile(P_k_q50, TOP_DECILE, na.rm = TRUE)
        cv_med <- median(CV_k, na.rm = TRUE)
        .SD[P_k_q50 >= qhi & CV_k < cv_med]
      }
    },
    by = .(model, hops_bin)
  ]
  if (!nrow(hi)) next

  dormant_loose  <- hi[any_exercised == FALSE | is.na(any_exercised)]
  dormant_strict <- hi[all_exercised == FALSE | is.na(all_exercised)]

  dormant_summary[[m]] <- hi[
    , .(n_top_decile_in_bin = .N), by = .(model, hops_bin)
  ][
    merge(dormant_loose[, .N, by = .(model, hops_bin)],
          dormant_strict[, .N, by = .(model, hops_bin)],
          by = c("model", "hops_bin"), all = TRUE,
          suffixes = c("_loose", "_strict")),
    on = c("model", "hops_bin"), nomatch = NA
  ][
    , .(model, hops_bin,
        n_top_decile = n_top_decile_in_bin,
        n_dormant_loose  = fcoalesce(N_loose,  0L),
        n_dormant_strict = fcoalesce(N_strict, 0L))
  ][
    , `:=`(frac_dormant_loose  = n_dormant_loose  / n_top_decile,
           frac_dormant_strict = n_dormant_strict / n_top_decile)
  ]

  if (!nrow(dormant_loose)) next

  per_path_nodes <- expand_paths_to_nodes(m, RISK_FORM_PRIMARY,
                                          paths_df = paths_df,
                                          node_df  = node_df)
  per_path_nodes <- per_path_nodes[path_id %in% dormant_loose$path_id]
  topk <- per_path_nodes[order(-risk_score),
                         head(.SD, TOP_K_FUNCTIONS),
                         by = .(model, path_id)]
  topk_collapsed <- topk[, .(
    top_functions = paste(name, collapse = " | "),
    top_files     = paste(unique(file), collapse = " | ")
  ), by = .(model, path_id)]
  out <- merge(dormant_loose[, .(model, path_id, hops_bin, hops, P_k_q50, CV_k)],
               topk_collapsed, by = c("model", "path_id"), all.x = TRUE)
  dormant_paths_out[[m]] <- out[order(hops_bin, -P_k_q50)]
}

if (length(dormant_summary)) {
  dormant_tbl <- rbindlist(dormant_summary, fill = TRUE)
  fwrite(dormant_tbl, file.path(syn_dir, "05_dormant_summary.csv"))
  fwrite(rbindlist(dormant_paths_out, fill = TRUE),
         file.path(syn_dir, "05_dormant_paths_table.csv"))
  print(dormant_tbl)
}
```

\newpage

# Cross-model synthesis

A single headline table per model collects the within-length effect
sizes, the rank correlation, the reverse predictive A on cyclomatic
complexity, the matched A, and the within-length dormancy. The
pre-registered decision rule ("framework predicts at runtime") is
evaluated last.

```{r synthesis}

# CROSS-MODEL SYNTHESIS ########################################################
################################################################################

headline <- data.table(model = reg$model)

# Coverage
cov_wide <- dcast(coverage_dt, model ~ stratum,
                  value.var = c("frac_any", "frac_all", "frac_edge"))
headline <- merge(headline, cov_wide, by = "model", all.x = TRUE)

# Rank correlation
if (exists("corr_results")) {
  headline <- merge(headline,
                    corr_results[, .(model,
                                     spearman_Pk_Eksum = spearman_sum,
                                     spearman_ci_lo    = rho_sum_ci_lo,
                                     spearman_ci_hi    = rho_sum_ci_hi)],
                    by = "model", all.x = TRUE)
}

# Reverse predictive (cyclomatic complexity only - the discriminating metric)
if (length(reverse_results)) {
  rev_cc <- rbindlist(reverse_results, fill = TRUE)[metric == "cyclomatic_complexity",
                                                    .(model, A_reverse_cc = effect_A)]
  headline <- merge(headline, rev_cc, by = "model", all.x = TRUE)
}

# Matched contrast (cyclomatic complexity only)
if (length(matched_results)) {
  m_cc <- rbindlist(matched_results, fill = TRUE)[metric == "cyclomatic_complexity",
                                                  .(model, A_matched_cc = effect_A,
                                                    n_matched_pairs = n_pairs)]
  headline <- merge(headline, m_cc, by = "model", all.x = TRUE)
}

# Dormancy (one row per model summarising the hops 3-4 bin)
if (length(dormant_summary)) {
  d_one <- rbindlist(dormant_summary, fill = TRUE)[hops_bin == "3-4",
                                                   .(model,
                                                     dormant_loose_3_4 = frac_dormant_loose)]
  headline <- merge(headline, d_one, by = "model", all.x = TRUE)
}

fwrite(headline, file.path(syn_dir, "06_cross_model_headline.csv"))
print(headline)

# Pre-registered decision rule per model
if (exists("corr_results") && length(reverse_results) && length(matched_results)) {
  cor_ok <- corr_results[, .(model,
                             spearman_pos = spearman_sum > 0 & rho_sum_ci_lo > 0)]
  rev_ok <- rbindlist(reverse_results, fill = TRUE)[,
    .(any_A_above_reverse = any(effect_A > A_THRESHOLD, na.rm = TRUE)),
    by = model]
  mat_ok <- rbindlist(matched_results, fill = TRUE)[,
    .(any_A_above_matched = any(effect_A > A_THRESHOLD, na.rm = TRUE)),
    by = model]
  decision <- merge(cor_ok, rev_ok, by = "model", all = TRUE)
  decision <- merge(decision, mat_ok, by = "model", all = TRUE)
  decision[, supports_framework :=
             spearman_pos & (any_A_above_reverse | any_A_above_matched)]
  fwrite(decision, file.path(syn_dir, "06_decision_rule.csv"))
  print(decision)
  n_support <- sum(decision$supports_framework, na.rm = TRUE)
  cat(sprintf("\n[decision rule] %d of %d available models meet the joint criterion. ",
              n_support, nrow(decision)))
  cat("Full pre-registered claim requires >= 3 of 4 target models.\n")
}
```

```{r synthesis_figure, fig.width=8, fig.height=8, fig.cap="Cross-model synthesis: coverage gap (top vs bottom decile, mean fraction of edges exercised), Spearman rho(P_k, E_k^sum) with bootstrap CI, and effect size A on cyclomatic complexity from Analyses 3 and 4. Dashed line at A = 0.55 is the pre-registered threshold."}

panels <- list()

p_cov <- ggplot(coverage_dt,
                aes(x = model, y = frac_edge, fill = stratum)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = color_strata,
                    labels = c("bottom decile", "top decile")) +
  labs(title = "Coverage gap (mean frac. of edges exercised)",
       x = NULL, y = NULL, fill = NULL) +
  theme_minimal(base_size = 10) + theme(legend.position = "bottom")
panels <- c(panels, list(p_cov))

if (exists("corr_results") && nrow(corr_results)) {
  p_corr <- ggplot(corr_results, aes(x = model, y = spearman_sum)) +
    geom_pointrange(aes(ymin = rho_sum_ci_lo, ymax = rho_sum_ci_hi)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    labs(title = expression("Spearman " ~ rho ~ "(P"[k] ~ ", E"[k]^{sum} ~ ")"),
         x = NULL, y = NULL) +
    theme_minimal(base_size = 10)
  panels <- c(panels, list(p_corr))
}

if (length(reverse_results) || length(matched_results)) {
  effect_dt <- rbind(
    if (length(reverse_results))
      rbindlist(reverse_results, fill = TRUE)[metric == "cyclomatic_complexity",
        .(model, analysis = "Analysis 3 (reverse)", A = effect_A)],
    if (length(matched_results))
      rbindlist(matched_results, fill = TRUE)[metric == "cyclomatic_complexity",
        .(model, analysis = "Analysis 4 (matched)", A = effect_A)],
    fill = TRUE
  )
  p_A <- ggplot(effect_dt, aes(x = model, y = A, fill = analysis)) +
    geom_col(position = position_dodge(0.7), width = 0.6) +
    geom_hline(yintercept = A_THRESHOLD, linetype = "dashed") +
    coord_cartesian(ylim = c(0, 1)) +
    labs(title = "Effect size A on cyclomatic complexity",
         x = NULL, y = NULL, fill = NULL) +
    theme_minimal(base_size = 10) + theme(legend.position = "bottom")
  panels <- c(panels, list(p_A))
}

cowplot::plot_grid(plotlist = panels, ncol = 1, align = "v")
```

\newpage

# Session information

```{r session_information, cache=FALSE}

# SESSION INFORMATION ##########################################################
################################################################################

sessionInfo()

cat("Machine:     "); print(get_cpu()$model_name)
cat("Num cores:   "); print(detectCores(logical = FALSE))
cat("Num threads: "); print(detectCores(logical = TRUE))
```
